Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-12 article.Rmd
Revised conclusion

The path from a dissertation to an AHR book review is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the AHR editors for review. As of August 2014 the journal’s website reminds readers that “the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue).” (http://www.indiana.edu/~ahrweb/book_review_guide.html - retrieved August 25 2014). Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More insidiously, women continue to face insidious gender discrimination.

@@

The American Historical Association also categorizes their reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The guiltiest parties, however, are also the two subject categories that publish the most book reviews in the AHR: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.

@@

One 2013 analysis of 2,500 recent history Ph.D.’s found that “gender played little role in employment patterns across particular professions and industries.” (http://historians.org/Documents/Many_Careers_of_History_PhDs_Final.pdf) Our own analysis reveals a more complicated reality. The flagship journal of professional historians continues to publish twice as many reviews of books by female authors as male authors. This disparity is put into even sharper relief when set against the relative parity achieved by women in producing history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the historical profession. Our program allows us to quantitatively measure the scope of the problem on a large scale, how it’s changed over time, and what remains to be done.

@@

Our contributions:

-     Broadly: gender studies + big data
-   	Temporal dimension makes it better suited to fit the needs of humanities research than contemporary approaches – both more accurate (historical) and more nuanced (supplying the gendered proportion of a name for different periods)
-   	Extensible – non US-names, older names. (future work)

The fundamental points are that we have a much larger corpus of names. More importantly the ability to vary the look up based on a specific date of birth or a specified range of dates. Making this much more useful for historians, but also for any research in any field, given that the normal human lifespan present in any data set means that names can vary as much as a century, and so everyone needs to be using a fundamentally historical method. (Take that, sociologists.)

>>>>

The path from a dissertation to an AHR book review is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the AHR editors for review. As of August 2014 the journal’s website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)."^[http://www.indiana.edu/~ahrweb/book_review_guide.html - retrieved August 25 2014]. Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More insidiously, women continue to face insidious gender discrimination.

@@

The American Historical Association also categorizes their reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The most uneven parties, however, are also the two subject categories that publish the most book reviews in the AHR: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.

@@

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries."^[http://historians.org/Documents/Many_Careers_of_History_PhDs_Final.pdf] Our own analysis reveals a more complicated reality. The flagship journal of professional historians continues to publish twice as many reviews of books by female authors as male authors. This disparity is put into even sharper relief when set against the relative parity achieved by women in producing history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the historical profession. Our program allows us to quantitatively measure the scope of the problem on a large scale, how it's changed over time, and what remains to be done.

@@

In conclusion, our algorith for predicting gender from first names combined with historical datasets in the Gender package provides a powerful new method for historians interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance historians always have to deal with the problem of scarcity in the sources. Our methods uses the abundant datasets to produce more information from scarce datasets. The temporal dimension of our work matches a historical question to a historical method, providing much improved over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make the predictions. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.
